Functory: A Distributed Computing Library \\ for Objective Caml \thanks{This research was partly supported by the French national project U3CAT (\emph{Unification of Critical C Code Analysis Techniques}, ANR-08-SEGI-021).}

### 2011-03-22 doc/tfp-2011/main.tex
proofreading intro and conclusion; adding nb_tasks

\begin{abstract}
  We present Functory, a distributed computing library for
  functional programming languages. The main features of this library
  include (1) a polymorphic API, (2) several implementations to
  adapt to different computing environments such as sequential,
  multi-core or network, and (3) a reliable fault-tolerance mechanism.
  This paper describes the motivation behind this work, as well as
  the design and implementation of the library. It also demonstrates
  the potential of the library using realistic experiments.
\end{abstract}

@@

This paper introduces \functory, a generic library for distributed
computing for a widely used functional programming language -
\Ocaml. This work was initially motivated by the computing needs that
exist in our own research team. Our applications include large-scale
deductive program verification, which amounts to checking the validity
of a large number of logical formulas using a variety of automated
theorem provers~\cite{filliatre07cav}. Our computing infrastructure
consists of a few powerful multi-core machines (typically 8 to 16
cores) and several desktop PCs (typically dual-core). However, for our
application needs, there is no library that helps in exploiting such a
computing infrastructure in our favorite functional programming
language.  Hence we designed and implemented such a library, which is
the subject of this paper.  Although this library is implemented in
\Ocaml, the implementation should be straightforward in any functional
programming language.

The distributed computing library presented in this paper is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of user-friendly APIs
that allows distributed execution of large-scale parallelizable
computations, very relevant to our application needs (and also
relevant to a variety of real-world applications). Further, the
distributed execution could be over multiple cores in the same machine
or over a network of machines. The most important features of our
library are the following:
\begin{itemize}
\item \emph{Genericity}: 
  it allows various patterns of polymorphic computations;
\item \emph{Simplicity}: switching between multiple cores on the same
  machine and a network of machines is as simple as changing a couple
  of lines of code;
\item \emph{Task distribution and fault-tolerance}: 
  it provides automatic task distribution and
  a robust fault-tolerance mechanisms, thereby
  relieving the user from implementing such routines.
\end{itemize}
The application domain of such a distributed computing library is manyfold. 
% It is worth noting that the library is not targeted 
\functory\ is not targeted at applications running on server farms,
crunching enormous amounts of data and hence it does not implement
techniques for data locality.  The library is very relevant for
research labs, to make use of available computing resources
efficiently. It serves a variety of users and a wide spectrum of
needs, from desktop PCs to networks of machines.  Typical applications
would involve executing a large number of computation expensive tasks
in a resource-optimal and time-efficient manner. This is also the case
in our research endeavours, that is validating thousands of
verification conditions using automated theorem provers, utilizing the
computing infrastructure to the maximum.

@@

The remainder of this section introduces our approach to distributed
computing in a functional programming setting.

\paragraph{Distributed Computing.}
A typical distributed computing library, as \functory, provides the
following (we borrow some terminology from Google's
MapReduce~\cite{mapreduce}):
\begin{itemize}
\item A notion of \emph{tasks} which denote atomic computations
  to be performed in a distributed manner; 
\item A set of processes (possibly executing on remote machines)
  called \emph{workers} that perform
  the tasks, producing results;
\item A single process called a \emph{master} which is in charge
  of distributing the tasks among the workers and managing results
  produced by the workers.
\end{itemize}
In addition to the above, distributed computing environments also
implement mechanisms for fault tolerance, efficient storage, and
distribution of tasks. This is required to handle network failures
that may occur, as well as to optimize the usage of machines in the
network. Another concern of importance is the transmission of messages
over the network. This requires efficient 
\emph{marshalling} of data, that is encoding and decoding of data 
for transmission over different computing environments.  It is desirable to
maintain architecture independence while transmitting marshalled data,
as machines in a distributed computing environment often run on
different hardware architectures and make use of different software
platforms. For example, machine word size or endianness may be different
across machines on the network.

@@

In the following, we discuss some related work in this domain and
elucidate how \functory\ differs from the other approaches.

@@

\JoCaml\ is one of the DFLs which provides communication primitives
(like channels) for facilitating transmission of
computations. However, it does not provide ready-made language
features for fault tolerance, which is indispensable in a distributed
setting. The user has to include code for fault tolerance, as already
demonstrated in some \JoCaml\
library~\cite{mandel2008}. ML5~\cite{ML5}, a variant of ML, is a
programming language for distributed computing, specialized for web
programming. It provides primitives for transferring control between
the client and the server, as well as low-level primitives for
marshalling the data. As in the case before, ML5 is a programming
language that offers primitives for code mobility, and the code for
distribution of computation and fault-tolerance has to be included by
the user.  On the contrary, ML5 implements type-safe marshalling and
\functory\ does not, though an existing type-safe marshalling library
could be used with \functory. Basically, the task will be transmitted
along with the corresponding type and on unmarshalling, it could be
matched for the correct type.
\\ \textbf{[Glasgow Distributed Haskell: TODO]}\\
CamlP3l~\cite{camlP3l} mixes the features of functional programming
with predefined patterns for parallel computation to offer a parallel
programming environment. Again, it is a programming language offering
primitives for distributing computation to parallel processes and also
to merge the results from parallel executions. Erlang~\cite{erlang}
is a programming language which has features for distribution and
fault tolerance. In particular, it has features for task distribution
and is more well-known for its rich error detection primitives and the
ability to support hot-swapping. The error detection primitives of
Erlang allow nodes to monitor processes in other nodes and also
facilitate automatic migration of tasks in failed nodes to recovered
or active nodes.

@@

There are several implementations of Google's Map/Reduce in functional
programming languages. But \of{Functory} was just inspired by Google's
Map/Reduce and is not exactly a Map/Reduce implementation. The
simplest difference comes from the very fact that \of{Functory} does not
operate on key/value pairs. 
% TODO other difference = absence of data locality
PlasmaMR~\cite{plasma} is an \Ocaml\ implementation of Google's
Map/Reduce on a distributed file system PlasmaFS. It is able to use
PlasmaFS to its advantage --- the ability of the file system to handle
large files and query functions that implement data locality to
optimize network traffic. However, PlasmaMR does not support fault
tolerance which is indespensible in any distributed computing
application.
% The Disco project~\cite{disco} is an implementation of
% Google's MapReduce in Erlang~\cite{erlang}.
Another Map/Reduce
implementation in \Ocaml\ is Yohann
Padioleau's~\cite{poor-man-mapreduce}.  It is built on
top of OCamlMPI~\cite{ocamlMPI}, while our approach uses a homemade
protocol for message passing. Currently, we have less flexibility
w.r.t. deployment of the user program than OCamlMPI; on the other
hand, we provide a more generic API together with fault tolerance.

@@

This section describes our API. We start from a generic distribution
function, \of{compute}, as sketched in the introduction. Then we
explain how this function is actually implemented in terms of
low-level primitives, which are also provided in our API.
The other way round, we also explain how the \of{compute} function can
be used to implement high-level distribution functions for map and
fold operations. Finally, we explain how these three APIs are
implemented in five different ways, according to five different deployment
scenarios.

@@

@@

The function \of{compute} above can actually be implemented in terms
of low-level primitives, such as adding a task, adding a worker,
performing one step of communication between master and workers, etc.
These primitives are provided in our API, such that the user can have
finer control over the execution of the distributed computation.
A type for distributed computations is introduced.
\begin{ocaml}
  type ('a, 'c) computation
\end{ocaml}
A computation is created with a function \of{create}, which accepts
the same \of{worker} and \of{master} as \of{compute}:
\begin{ocaml}
  val create : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> ('a, 'c) computation
\end{ocaml}
Contrary to \of{compute}, it takes no list of tasks and returns
immediately. Tasks can be added later using the following function:
\begin{ocaml}
  val add_task : ('a, 'c) computation -> 'a * 'c -> unit
\end{ocaml}
A function is provided to perform \emph{one step} of a given
computation:
\begin{ocaml}
  val one_step : ('a, 'c) computation -> unit
\end{ocaml}
Calling this function results in one exchange of messages between
master and workers: task assignments to workers, results returned to
the master, etc. A few other functions are provided, such as
\of{status} to query the status of a computation, \of{clear} to remove
all tasks, etc.

@@

\subsection{Functory API at a Glance} % ???

From  the description above, it is clear that our library provides
three different APIs, for five different deployment scenarios,
resulting in fifteen possible ways of using it.
This is summarized in the following matrix.
\begin{center}
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & & & \multicolumn{3}{c|}{Network} \\
    & Seq. & Cores & Same & Poly & Mono \\\hline\hline
    low-level API & & & & & \\\hline
    \of{compute}  & & & & & \\\hline
    high-level API& & & & & \\\hline
  \end{tabular}
\end{center}
From the implementation point of view, there is almost no code
duplication. We are using \Ocaml\ functors to derive the second line
from the first, as well as the third from the second, irrespective of
the deployment scenario.

@@

In this paper, we presented a distributed programming library for the
functional programming language \Ocaml. The main features are the
genericity of the interface, which makes use of polymorphic
higher-order functions, and the ability to easily switch between
sequential, multi-core, and network implementations. In particular,
\functory\ allows to use the same executable for master and workers,
which makes the deployment of small programs immediate --- master and
workers being only distinguished by an environment
variable. \functory\ also allows master and workers to be completely
different programs, which is ideal for large scale deployment.
Another distinguishing feature of our library is a robust
fault-tolerance mechanism which relieves the user of cumbersome
implementation details.  Yet another interesting feature of the
library is the ability to add workers dynamically. Finally, \functory\
also allows to cascade several distributed computations inside the
same program.  \functory also implements two levels of API with
different granularities which enables the user to write interactive
programs.

@@

  Currently, without any information about the tasks, the scheduling
  is completely arbitrary. In both \of{Cores} and \of{Network}
  modules, we use traditional queues for the pending tasks; in
  particular, new tasks produced by the master are appended at the end
  of the queue.

% \item 
%   Another interesting feature could be the ability
%   to add or remove machines dynamically. Currently, our library assumes
%   that the list of machines to be used is given \emph{a priori}, as
%   part of the code. An alternative would be to read machine names from
%   a file, watched periodically by the master.
\item
  Our library provides limited support for displaying real-time
  information about computations and communications. Processing and storing
  information about workers and tasks locally in the master is straightforward; 
  monitoring it in real-time could be done using
  \textsf{Ocamlviz}~\cite{ocamlviz}. 
\item 
  One very nice feature of Google's MapReduce is the possibility to
  use redundantly several idle workers on the same tasks
  for speedup when reaching the end of computation.
  Since we already have the fault tolerance implemented, this
  optimization should be straightforward to add to our library.
\end{itemize}
We intend to enrich our library with all above features.

>>>>

\begin{abstract}
  We present Functory, a distributed computing library for
  Objective Caml. The main features of this library
  include (1) a polymorphic API, (2) several implementations to
  adapt to different deployment scenarios such as sequential,
  multi-core or network, and (3) a reliable fault-tolerance mechanism.
  This paper describes the motivation behind this work, as well as
  the design and implementation of the library. It also demonstrates
  the potential of the library using realistic experiments.
\end{abstract}

@@

This paper introduces \functory, a generic library for distributed
computing for a widely used functional programming language,
Objective Caml (\Ocaml\ for short). 
This work was initially motivated by the computing needs that
exist in our own research team. Our applications include large-scale
deductive program verification, which amounts to checking the validity
of a large number of logical formulas using a variety of automated
theorem provers~\cite{filliatre07cav}. Our computing infrastructure
consists of a few powerful multi-core machines (typically 8 to 16
cores) and several desktop PCs (typically dual-core). However, for our
application needs, there is no library that helps in exploiting such a
computing infrastructure in our favorite functional programming
language.  Hence we designed and implemented such a library, which is
the subject of this paper.  
% Although this library is implemented in
% \Ocaml, the implementation should be straightforward in any functional
% programming language.

The distributed computing library presented in this paper is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of user-friendly APIs
that allows distributed execution of large-scale parallelizable
computations, very relevant to our application needs (and also
relevant to a variety of real-world applications). Further, the
distributed execution could be over multiple cores in the same machine
or over a network of machines. The most important features of our
library are the following:
\begin{itemize}
\item \emph{Genericity}: 
  it allows various patterns of polymorphic computations;
\item \emph{Simplicity}: switching between multiple cores on the same
  machine and a network of machines is as simple as changing a couple
  of lines of code;
\item \emph{Task distribution and fault-tolerance}: 
  it provides automatic task distribution and
  a robust fault-tolerance mechanisms, thereby
  relieving the user from implementing such routines.
\end{itemize}
The application domain of such a distributed computing library is manyfold. 
% It is worth noting that the library is not targeted 
\functory\ is not targeted at applications running on server farms,
crunching enormous amounts of data, such as Google's
MapReduce~\cite{mapreduce}.  The library is very relevant for
research labs, to make use of available computing resources
efficiently. It serves a variety of users and a wide spectrum of
needs, from desktop PCs to networks of machines.  Typical applications
would involve executing a large number of computation expensive tasks
in a resource-optimal and time-efficient manner. This is also the case
in our research endeavours, that is validating thousands of
verification conditions using automated theorem provers, utilizing the
computing infrastructure to the maximum.

@@

In the following, we introduce our approach to distributed
computing in a functional programming setting and distinguish it from
related work.

\paragraph{Distributed Computing.}
A typical distributed computing library, as \functory, provides the
following (we borrow some terminology from Google's
MapReduce):
\begin{itemize}
\item A notion of \emph{tasks} which denote atomic computations
  to be performed in a distributed manner; 
\item A set of processes (possibly executing on remote machines)
  called \emph{workers} that perform
  the tasks, producing results;
\item A single process called a \emph{master} which is in charge
  of distributing the tasks among the workers and managing results
  produced by the workers.
\end{itemize}
In addition to the above, distributed computing environments also
implement mechanisms for fault tolerance, efficient storage, and
distribution of tasks. This is required to handle network failures
that may occur, as well as to optimize the usage of machines in the
network. Another concern of importance is the transmission of messages
over the network. This requires efficient 
\emph{marshalling} of data, that is encoding and decoding of data 
for transmission over different computing environments.  It is desirable to
maintain architecture independence while transmitting marshalled data,
as machines in a distributed computing environment often run on
different hardware architectures and make use of different software
platforms. For example, machine word size or endianness may be different
across machines on the network.

@@

% \medskip

% In the following, we discuss some related work in this domain and
% elucidate how \functory\ differs from the other approaches.

@@

\JoCaml\ is one of the DFLs which provides communication primitives
(like channels) for facilitating transmission of
computations. However, it does not provide ready-made language
features for fault tolerance, which is indispensable in a distributed
setting. The user has to include code for fault tolerance, as already
demonstrated in some \JoCaml\
library~\cite{mandel2008}. ML5~\cite{ML5}, a variant of ML, is a
programming language for distributed computing, specialized for web
programming. It provides primitives for transferring control between
the client and the server, as well as low-level primitives for
marshalling the data. As in the case before, ML5 is a programming
language that offers primitives for code mobility, and the code for
distribution of computation and fault-tolerance has to be included by
the user.  ML5 implements type-safe marshalling and
\functory\ does not, though an existing type-safe marshalling library
could be used with \functory. 
% Basically, the task will be transmitted
% along with the corresponding type and on unmarshalling, it could be
% matched for the correct type.
\\ \textbf{[Glasgow Distributed Haskell: TODO]}\\
CamlP3l~\cite{camlP3l} mixes the features of functional programming
with predefined patterns for parallel computation to offer a parallel
programming environment. Again, it is a programming language offering
primitives for distributing computation to parallel processes and also
to merge the results from parallel executions. Erlang~\cite{erlang}
is a programming language which has features for distribution and
fault tolerance. In particular, it has features for task distribution
and is more well-known for its rich error detection primitives and the
ability to support hot-swapping. The error detection primitives of
Erlang allow nodes to monitor processes in other nodes and also
facilitate automatic migration of tasks in failed nodes to recovered
or active nodes.

@@

There are several implementations of Google's MapReduce in functional
programming languages. But \of{Functory} was just inspired by Google's
MapReduce and is not exactly a MapReduce implementation. The
simplest difference comes from the very fact that \of{Functory} does not
operate on key/value pairs. 
% TODO other difference = absence of data locality
PlasmaMR~\cite{plasma} is an \Ocaml\ implementation of Google's
MapReduce on a distributed file system PlasmaFS. It is able to use
PlasmaFS to its advantage --- the ability of the file system to handle
large files and query functions that implement data locality to
optimize network traffic. However, PlasmaMR does not support fault
tolerance which is indespensible in any distributed computing
application.
% The Disco project~\cite{disco} is an implementation of
% Google's MapReduce in Erlang~\cite{erlang}.
Another MapReduce
implementation in \Ocaml\ is Yohann
Padioleau's~\cite{poor-man-mapreduce}.  It is built on
top of OCamlMPI~\cite{ocamlMPI}, while our approach uses a homemade
protocol for message passing. Currently, we have less flexibility
w.r.t. deployment of the user program than OCamlMPI; on the other
hand, we provide a more generic API together with fault tolerance.

@@

This section describes our API. 
We start from a simple API which is reduced to a single higher-order
polymorphic function.
Then we explain how this function is actually implemented in terms of
low-level primitives, which are also provided in our API.
The other way round, we also explain how the same function can
be used to implement high-level distribution functions for map and
fold operations. Finally, we explain how our API is
implemented in five different ways, according to five different deployment
scenarios.

@@

% TODO if we have space for it: small example to show the use of 'c
% 
%   let array_map f a =
%     let h = Hashtbl.create 17 in
%     let master (_, i) y = Hashtbl.add h i y in
%     let tasks = ref [] in
%     for i = 0 to Array.length a - 1 do tasks := (a.(i), i) :: !tasks done;
%     compute ~worker:f ~master !tasks;
%     Array.init n (fun i -> Hashtbl.find h i)

@@

The function \of{compute} above can actually be implemented in terms
of low-level primitives, such as adding a task, adding a worker,
performing one step of communication between master and workers, etc.
These primitives are provided in our API, such that the user can
interact with the execution of the distributed computation.  For
instance, a monitoring-like application can use these primitives to
allow observation and modification of resources (tasks, workers)
during the course of a computation.

A type for distributed computations is introduced:
\begin{ocaml}
  type ('a, 'c) computation
\end{ocaml}
A computation is created with a function \of{create}, which accepts
the same \of{worker} and \of{master} as \of{compute}:
\begin{ocaml}
  val create : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> ('a, 'c) computation
\end{ocaml}
Contrary to \of{compute}, it takes no list of tasks and returns
immediately. Tasks can be added later using the following function:
\begin{ocaml}
  val add_task : ('a, 'c) computation -> 'a * 'c -> unit
\end{ocaml}
A function is provided to perform \emph{one step} of a given
computation:
\begin{ocaml}
  val one_step : ('a, 'c) computation -> unit
\end{ocaml}
Calling this function results in one exchange of messages between
master and workers: task assignments to workers, results returned to
the master, etc. A few other functions are provided, such as
\of{status} to query the status of a computation, \of{clear} to remove
all tasks, etc.

@@

\subsection{Several APIs in One}

From  the description above, it is clear that our library provides
several APIs of different granularities and different
deployment scenarios. 
Most combinations are meaningful, resulting in
thirteen possible different ways of using our library.
% This is summarized in the following matrix.
% \begin{center}
%   \begin{tabular}{|l|c|c|c|c|c|}
%     \hline
%     & & & \multicolumn{3}{c|}{Network} \\
%     & Seq. & Cores & Same & Poly & Mono \\\hline\hline
%     low-level API & & & & & \\\hline
%     \of{compute}  & & & & & \\\hline
%     high-level API& & & & & \\\hline
%   \end{tabular}
% \end{center}
From the implementation point of view, there is almost no code
duplication. We are using \Ocaml\ functors to derive specific
implementations from generic ones.

@@

In this paper, we presented a distributed programming library for
\Ocaml. The main features are the
genericity of the interface, which makes use of polymorphic
higher-order functions, and the ability to easily switch between
sequential, multi-core, and network implementations. In particular,
\functory\ allows to use the same executable for master and workers,
which makes the deployment of small programs immediate --- master and
workers being only distinguished by an environment
variable. \functory\ also allows master and workers to be completely
different programs, which is ideal for large scale deployment.
Another distinguishing feature of our library is a robust
fault-tolerance mechanism which relieves the user of cumbersome
implementation details.  Yet another interesting feature of the
library is the ability to add workers dynamically. \functory\
also allows to cascade several distributed computations inside the
same program.  Finally, the low-level API of \functory\ 
can be used to write interactive programs.

@@

  Currently, without any information about the tasks, the scheduling
  is completely arbitrary. In both \of{Cores} and \of{Network}
  modules, we use traditional queues for the pending tasks; in
  particular, new tasks produced by the master are appended to the end
  of the queue.

% \item 
%   Another interesting feature could be the ability
%   to add or remove machines dynamically. Currently, our library assumes
%   that the list of machines to be used is given \emph{a priori}, as
%   part of the code. An alternative would be to read machine names from
%   a file, watched periodically by the master.
\item
  Our library provides limited support for retrieving real-time
  information about computations and communications. Processing and storing
  information about workers and tasks locally in the master is
  straightforward.

\item 
  One very nice feature of Google's MapReduce is the possibility to
  use redundantly several idle workers on the same tasks
  for speedup when reaching the end of computation.
  Since we already have the fault tolerance implemented, this
  optimization should be straightforward to add to our library.
\end{itemize}
We intend to enrich our library with all above features.
